Coordinations are structures which consist of multiple elements, such that neither of the elements governs any of the others. Those elements are conjuncts and they can be connected with punctuation and with conjunctions such as \textsl{and}, \textsl{or}, \textsl{as well as} and others. An example of a coordination is provided in (\ref{ex:coord}), with a conjunction \textsl{and} and conjuncts: \textsl{some apples}, \textsl{the oranges your mother gave you}.

\begin{exe}
    \ex\label{ex:coord}
    Bring [[some apples] and [the oranges your mother gave you]]
\end{exe}

This work aims to investigate how coordinations should be interpreted in certain syntactic theories, specifically in dependency grammars. In dependency grammars the structure of a sentence is represented by a dependency tree, in which almost every word depends on another one -- that word is its governor. In (\ref{ex:depTree}) there is an example of a sentence with a dependency annotation. In this sentence the shorter conjunct is placed as the left one, which was observed to be common in the English language.

\begin{exe}
	\ex\label{ex:depTree}
	\begin{dependency}[baseline=-\the\dimexpr\fontdimen22\textfont2\relax]
	\begin{deptext}
	The\& history\& book\& on\& the\& shelf\& is\& always\& repeating\& itself\\ 
	\end{deptext}
	\deproot{7}{root}
	\depedge[edge height=1.5cm]{7}{3}{subj}
	\depedge{7}{8}{mod}
	\depedge{7}{9}{comp:aux}
	\depedge{3}{1}{det}
	\depedge{3}{2}{compound}
	\depedge{3}{4}{mod}
	\depedge{4}{6}{comp:obj}
	\depedge{6}{5}{det}
	\depedge{9}{10}{comp:obj} %wg parsera tu powinno być udep@npmod BŁĘDNIE
	\end{dependency}
\end{exe}

\cite{prz:woz:23} have conducted a study in which they provided an argument for certain approaches to annotating coordinations in dependency grammars. To do so, they checked whether coordinate structures were formed differently depending on the position of the governor of the structure. They found that in coordinations with a governor on the left (as in (\ref{ex:GL}) with the governor \textsl{bring}) it is more likely that the shorter conjunct will be also placed on the left. This means that people are more likely to form sentence (\ref{ex:GLSL}) rather than sentence (\ref{ex:GLSR}). Similarly, if the coordination has no governor (as in (\ref{ex:GN})), the shorter conjunct is more likely to be on the left (as in (\ref{ex:GNSL})) than on the right (as in (\ref{ex:GNSR})).

\begin{exe}

\ex\label{ex:GL}
\begin{xlist}
\ex\label{ex:GLSL}
\textsl{Bring} [[some apples] and [the oranges your mother gave you]]

\ex\label{ex:GLSR}
\textsl{Bring} [[the oranges your mother gave you] and [some apples]]
\end{xlist}

\ex\label{ex:GN}
\begin{xlist}
\ex\label{ex:GNSL}
[[Buy some apples] or [steal as many oranges as you can hold]]

\ex\label{ex:GNSR}
[[Steal as many oranges as you can hold] or [buy some apples]]
\end{xlist}

\end{exe}

However in sentences with a governor on the right, the shorter conjunct is more likely to appear as the right one, which means that sentence (\ref{ex:GRSR}) is more likely to appear in the English language than sentence (\ref{ex:GRSL}).

\begin{exe}

\ex\label{ex:GR}
\begin{xlist}
\ex\label{ex:GRSL}
[[An apple] and [three long orange peels]] \textsl{fell} out of the bag

\ex\label{ex:GRSR}
[[Three long orange peels] and [an apple]] \textsl{fell} out of the bag
\end{xlist}

\end{exe}

Those findings were based on the Penn Treebank, which is a corpus of texts collected from the Wall Street Journal with manually added syntactic annotation. One advantage of such a resource is high quality, reliable annotation, but since creating manual annotation requires time, the corpus is small. Additionally, the corpus contains only texts from the Wall Street Journal, therefore it is not stylistically diverse. 

Another paper by \cite{prz:etal:24} describes an attempt at replicating the results found in \cite{prz:woz:23} on a larger and more diverse corpus, namely the Corpus of Contemporary American English. Downside to using this corpus is that there is no syntactic annotation, therefore it had to be added automatically using a parser. This unfortunately lowered the quality of data, compared to the manually annotated Penn Treebank. After conducting an evaluation, \cite{prz:etal:24} found only around half of the sampled coordinate structures to be correctly extracted from the text. 

The dependency annotation in \cite{prz:etal:24} was made according to a scheme called Universal Dependencies. \cite{tuo:prz:lac:21} show that some automatic parsers can perform better when using another scheme called Surface-syntactic Universal Dependencies. Additionally, the structural differences between those two schemes allow for more precise heuristics for finding coordinations in sentences parsed according to the surface-syntactic scheme.

The main goal of the current study is to replicate the aforementioned studies -- this provides the first two hypotheses that are tested here: 1) placement of the shorter conjunct within a coordination depends on the position of the governor and 2) the tendency for the shorter conjunct to appear closer to the governor is more prominent with bigger length differences between conjuncts. The third hypothesis is a result of an effort to increase the quality of the automatically annotated and extacted data. It states that the parsing accuracy is higher when text is parsed according to Surface-syntactic Universal Dependencies, rather than the original Universal Dependencies scheme.

The structure of this thesis is the following: Chapter 2 explains the necessary theoretical background. Chapter 3 describes how the corpus data was prepared for analysis -- from training the parser to extracting the information about coordinate structures. Chapter 4 presents the analysis of the data, which is discussed in Chapter 5. Finally Chapter 6 covers the limitations of this research. 